---
title: "Cleaning The Data"
execute: 
  output: false
---

Some sort of introductory text to what cleaning data even means.

## 1. Setup

First, we need to load the packages that are going to help facilitate the data analysis. You don't need to worry about what each package does, but this is an important setup step that allows us to use specific functions later on. I always include these three at the top of every document. Copy and paste the code below and "run" it by hitting the green play button in the top right of your code chunk.

```{r}
#| label: setup
# setup 
library(tidyverse)
library(janitor)
library(lubridate)
```

You'll probably see some sort of message pop up after you run the code; as long as there's no red "error" warning, you're fine to move on.

## 2. Importing the data from LegiScan

Now you need to take the data that we downloaded to your computer and import it into this R file. It sounds complicated, but we can actually do this with a simple line of code.

If you open the files we downloaded from LegiScan in the previous step, you'll see it actually includes seven different data tables. For the purposes of this project, we're only going to be working with three of those: "bills," "sponsors," and "people." Copy and run the code below to import those three tables into this file.

```{r}
#| label: import bills
 bills <- read_csv("data-raw/tx-bills-all/2025-2026_89th_Legislature/csv/bills.csv")
 
 bills
```

```{r}
#| label: import sponsors
sponsors <- read_csv("data-raw/tx-bills-all/2025-2026_89th_Legislature/csv/sponsors.csv")

sponsors
```

```{r}
#| label: import people
lawmakers <- read_csv("data-raw/tx-bills-all/2025-2026_89th_Legislature/csv/people.csv")

lawmakers
```

## 3. "Cleaning" those tables

Oftentimes, data tables aren't in the best shape when you download them. There might be rows or cells that are missing entries, or there may be columns that aren't necessary for the analysis you want to do. This is where cleaning comes in: you can use code to adjust the table/remove data you don't need so the final table is much easier to work with.

### 3.1 Bills

Let's clean up each of the data tables one at a time. We'll start with our "bills" data.

##### A) View the raw data

Before we start cleaning up the tables, let's take another look at what they look like untouched. Run this code to view the table you just imported.

```{r}
#| label: view bill data
bills
```

add description of what we're looking at here

##### B) Remove unnecessary columns and rename columns

```{r}
#| label: clean bill data 
bills_clean <- bills |> 
   select(-session_id, -status, -status_desc, -title, -committee_id, -committee, -last_action_date, -last_action, -url, -state_link) |> 
  rename(date_filed = status_date)

bills_clean 
```

Describe what we're looking at now.

### 3.2 Lawmakers

##### A) View the raw data

```{r}
#| label: view lawmaker data 
lawmakers 
```

Description of what we've got here.

##### B) Remove unnecessary columns and rename columns

```{r}
#| label: clean lawmaker data 
lawmakers_clean <- lawmakers |> 
  select(people_id, name, party, role, district) 

lawmakers_clean
```

Description of what we're working with now.

### 3.3 Sponsors

##### A) View the raw data

```{r}
#| label: view sponsor data 
sponsors
```

##### B) Remove unnecessary columns and rename columns

```{r}
#| label: clean sponsor data 

sponsors |> 
  select(-position)

```

Restate all the changes made.

## 4. Merge the tables together

Now that we've selected the columns we want to work with in each of the three data tables, we need to merge everything together into one big table that we'll use for our analysis. You'll notice there are a couple of repeat columns in the tables, such as people_id and bill_id. We'll want to make sure those only appear once in our big table. Let's do this in two steps: first, copy the code below to merge the lawmaker and sponsor tables together.

```{r}
lawmaker_sponsor <- merge(lawmakers_clean, sponsors) |> 
  select(-position)

lawmaker_sponsor
```

Now lets add the description column and date_filed column from the bills table.

```{r}
final_table <- lawmaker_sponsor |> 
  left_join(bills_clean, by = join_by(bill_id)) 

final_table
```

Great! You should have one data table with the following information:

-   name = lawmaker's name

-   people_id = the unique number associated with each lawmaker

-   party = political party for each lawmaker

-   role = whether the lawmaker is a senator or a representative

-   district = the congressional district associated with each lawmaker

-   bill_id = the unique number associated with a specific bill filed this session

-   date_filed = the date that bill was introduced

-   description = explains what a specific bill aims to do

We'll be able to quickly sort through all of this information and identify key details about the agendas of Texas lawmakers in the current legislative session. Let's save this table before we move forward with our analysis.

## 5. Saving the clean table

Add instructions for creating a data-processed file.

```{r}
final_table |> 
  write_rds("data-processed/bills-89.rds")
```

```{r}
final_table |> 
  write_csv("data-processed/bills-89.csv")
```

In the next chapter of this toolkit, we'll brain storm some potential questions about the session and use the data to find some answers.
